Learning by Teaching a Robot:\\The Case of Handwriting

### 2016-03-16 main.tex
Double column layout
Make the paper shorter

%\documentclass[twocolumn]{article}
\documentclass{article}

@@

\subsection{Learning by Teaching Handwriting}

@@

Robots have been used as teachers or social partners to promote children's
learning in a range of contexts, most commonly related to language
skills~\cite{han2010robot}, and less often to physical skills (such as
calligraphy~\cite{Matsui2013}). Looking at the converse (humans \emph{teaching}
robots), Werfel notes in~\cite{Werfel2014} that most of the work focuses on the
robot's benefits (in terms of language~\cite{Saunders2010} or
physical~\cite{Mulling2013} skills, for example) rather than the learning
experienced by the human tutor themselves.  Our work concentrates on this latter
aspect: by demonstrating handwriting to a robot, we aim at improving the
\emph{child's} performance. Note that our work must be distinguished from
``learning from demonstration'' approaches to robots learning physical skills,
as the agent we present is only simulating poor fine motor skills for interaction
purposes.

@@

\subsection{Agency and Commitment}

@@

For these two reasons, our approach is to be distinguished from previous works
in educational robotics. Most of these do not consider the agency induced by the
robot beyond its motivational aspect (playing with an interactive, partially
autonomous device naturally induces some form of anthropomorphizing, which leads
to some level of projected agency, and contributes to the overall excitement -- at
least, on the short-term, before the novelty effect vanishes).  In our case, the
role of agency is stronger: it induces meta-cognition (``I am interacting with
an agent, so I need to reflect on how to best teach him'') which is beneficial
for the learning process; it also induces a protégé effect (``I want my
robot-agent to succeed!'') which supports the commitment of the child into the
interaction, also for longer periods of time.

Building on these socio-cognitive mechanisms, our intent is therefore to design
a robotic system that would effectively support handwriting remediation in an
original way. By getting children to teach a robotic agent how to write, those
children would both practise without knowing it and recover self-confidence and
self-esteem by supporting a worse-than-themselves robotic ``student''.

\subsection{Research Questions}

These initial considerations can be turned into a set of research questions.

\paragraph{Technical Feasibility} The technical feasibility of our system (and
its actual implementation) is the first question -- actually a prerequisite --
that we want to address. The robotic system needs to be able to produce
handwriting in a believable manner; the legibility of its handwriting (\ie the
apparent skill of the robot) needs to be easy to parametrise and must evolve
over time in a controlled manner; the robot needs to acquire, recognize and
learn from handwriting demonstrations that may be hardly legible (due to the
impairments of our target children); the design and implementation of the
robot's behaviors must effectively sustain and support a complex and long-term
interaction with children; and finally, in contrast to previous work, we aim at
building a fully autonomous agent that can be deployed and used by non-experts
(teachers, therapists). Taken together, these technical objectives are
ambitious and concern a range of different technical subcomponents.

\paragraph{Acceptance} A second research question that we study here relates to
the acceptance of such a system in educative and medical contexts. Remediation
to cognitive deficits, even mild ones, is a sensitive question for the
caretakers, all the more so when working with children. Introducing a
robot-based system in such situations is not self-evident, especially as the
mentor-protégé relationship further relies on building strong affective and
cognitive bonds between the child and a machine (here, the robot is not simply
experienced as a passive tool, as would be an educational computer program for
instance: it instead plays a potentially disruptive role). By establishing
collaborations with teachers, parents and therapists, we aim however at going
down to the practitioners' field level and explore the conditions for such a
robot-based system to be accepted.

\paragraph{Engagement} We have already mentioned that handwriting remediation
requires exercising over long periods of time. In terms of human-robot
interaction, this translates into a third research question: how to sustain
several hours of interaction while performing a task (handwriting) which is
essentially perceived as a boring routine? Sustained interaction is usually
studied in term of \emph{engagement}: how to engage, and thereafter sustain the
engagement, of a child into an interaction with a robot? We hypothesize that the
\emph{learning by teaching} paradigm may be an effective lever to get the child
to commit in helping the robot over an extended period of time. This needs to be
evidenced -- which incidentally calls for setting up and carrying long-term
child-robot experiments.

\paragraph{Remediation Efficacy} Finally, the last, key, research question is: is
our approach effective in helping children to recover from handwriting deficits?
Answering this question in a broad, general way is unlikely as {\it 1)} the
observed deficits vary a lot between the children (no single metric) and have
complex, compound causes, and, {\it 2)} ``learning how to write'' is not a
monolithic, linear skill that could be easily rated from 0 to 1.  However, by
crossing qualitative assessments made by the caretakers with quantitative
measures of visual similarity between the letters drawn by the children and some
reference templates, we aim at assessing the impact of our system on the
children' handwriting, if only in terms of trends.

\vspace{2em}
The following sections explore and attempt to answer these
questions. We first provide an
overview of the robotic system and the interaction it induces
in Section~\ref{impl}. We present there
the machine-learning techniques that allow the robot to learn from the children,
as well as the actual implementation of the system on a {\sc nao} robot.

@@

Figure~\ref{experimental_setup} illustrates our general experimental setup: a
face-to-face child-robot interaction with an autonomous Aldebran's {\sc nao}
robot.

A tactile tablet (with a custom application) is used for both the robot and the
child to write. During a typical round, the child requests the robot to write
something (a single letter, a number or a full word), and pushes the tablet
towards the robot. The robot writes on the tablet by gesturing the writing in
the air,  the letters being actually drawn by the tablet application itself. The
child then pulls back the tablet, corrects the robot's attempt by writing
him/herself on top of or next to the robot's writing (see Figure~\ref{fig:diego}),
and ``sends'' his/her demonstration to the robot by pressing a small button on
the tablet. The robot ``learns'' from this demonstration and tries again.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.6\columnwidth]{experimental_setup}
    \caption{\small Our experimental setup: face-to-face interaction with a {\sc
        nao} robot.  The robot writes on the tactile tablet, the child then
        corrects the robot by directly overwriting its letters on the tablet
        with a stylus. An adult (either a therapist or an experimenter,
        depending on the studies), remains next to the child to guide the work
        (prompting, turn taking, etc.). For some studies, a second tablet and an
        additional camera (dashed) are employed.}

@@

We detail each of these in the following sections.

@@

The main insight for both the generation and the learning of letters is to
reason about the shape of letters in their eigenspace, instead of the natural
cartesian space. The eigenspace of each letter is spanned by the first $n$
eigenvectors (in our experiments, $3 < n < 6$) of the covariance matrix
generated from a standard dataset of adult letters (the UJI Pen Characters 2
dataset~\cite{Llorens2008}).  This procedure, based on a Principle Component
Analysis (PCA), is explained in details in a previously published
article~\cite{hood2015when}.

New letters can be generated by varying the weightings in linear combinations of
these eigenvectors, with distortions that are actually plausible handwriting errors: 
they are exaggerations of variations of writing styles that naturally occur amongst
adult writers.  Figure~\ref{fig:sampleLetters} shows examples of deformed ``g''
generated with such a technique.

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{cowriter-g}
    \caption{\small \label{fig:sampleLetters} \textbf{Generating bad letters}:
        effect of independently varying the weightings (columns) of the first five eigenvectors 
        (rows) of the shape model of ``g''. Examples A, B, C, D illustrate how the
        PCA-based approach allows for the \emph{automatic} generation of letters whose
        errors can be \emph{semantically} interpreted, \eg A has too large a bottom
        loop, B has a wide top loop, the bottom loop of C is not correctly
        closed, the top loop of D is not closed, etc.}

@@

The same shape model can also be used to \emph{classify} demonstrations,
\emph{assess} their quality and \emph{learn} from them. Figure~\ref{fig:h} shows
for example nine allographs of ``h'' written by a 6 years old child, along with
a reference letter. By projecting each of the
demonstrations onto the eigenspace of ``h'' (Figure~\ref{fig:h}b), we observe
that:

@@

\begin{figure}[ht!]
    \centering
    \subfigure[Nine allographs of the cursive ``h'', next to the reference]{
        \includegraphics[width=0.45\columnwidth]{h}
    }
    \subfigure[The same samples, normalized and projected in the eigenspace spanned by the
        first 3 eigenvectors: clusters arise, that actually match writing styles.]{
        \includegraphics[width=0.45\columnwidth]{eigenspace-uniformization}
    }

    \caption{\small Projecting demonstrated letters onto the eigenspace
    generated from the reference dataset effectively clusters the samples
    according to their topological similarity. Allographs that are similar to
    the reference are close to it in the eigenspace.}
    \label{fig:h}
\end{figure}

@@

\begin{figure}[ht!]
\centering

\resizebox{0.7\linewidth}{!}{%

\begin{tikzpicture}[
    >=latex,
    node distance=2cm,
    every edge/.style={draw, very thick},
    redarrow/.style={draw,red, text=black},
    greenarrow/.style={draw,GreenYellow,text=black},
    yellowarrow/.style={draw,BurntOrange,text=black},
    cmpt/.style={draw, align=center, rounded corners, inner sep=5pt, font=\sf, fill=black!20},
    label/.style={midway, align=left, font=\scriptsize\sf, fill=white, above,opacity=0,text opacity=1}]

    \node at (0,0) (laptop) {\includegraphics[width=2cm]{laptop}};
    \node[below right=2 of laptop] (nao) {\includegraphics[width=2cm]{nao}};
    \node[below left=2 of laptop] (tablet) {\includegraphics[width=2cm]{tablet+stylus}};
    \node[above=2 of laptop] (selection) {\includegraphics[width=2cm]{selection_tablet}};

    \node[draw,above right=2 of laptop,anchor=north west,text width=4cm] (processes)
    {\sf\scriptsize machine-learning, \\letters/gestures
    generation, \\interaction supervision};
    \path (laptop) edge[dashed] (processes);

    \path (nao) edge [->,redarrow, bend left] node[label, auto] {robot state} (laptop);
    \path (laptop) edge [->,greenarrow, bend left] node[label, auto] {writing gestures} (nao);

    \path (tablet) edge [->,redarrow, bend left] node[label, auto] {demonstrations,\\turn taking} (laptop);
    \path (laptop) edge [->,redarrow, bend left] node[label,
    auto=right,align=right]
    {path of\\ letters to display} (tablet);

    \path (selection) edge [->,redarrow] node[label, auto=right] {letter/word to write} (laptop);

    \path (-5, 2) edge [->, redarrow] node[label] {ROS} ++(1, 0);
    \path (-5, 2.6) edge [->, greenarrow] node[label] {NaoQI} ++(1, 0);

\end{tikzpicture}
}

\caption{\small \textbf{Overview of the system}. In total, the system runs about 10 ROS nodes,
    distributed over the robot itself, a central laptop and Android tablets.}

    \label{fig:archi}
\end{figure}

The overall architecture of the system (Figure~\ref{fig:archi}) is therefore
spread over several devices: the {\sc nao} robot itself, that we address via
both a ROS API\footnote{The ROS documentation for {\sc nao} is available at
\url{http://wiki.ros.org/nao}.} and the Aldebaran-provided NaoQI API, one to
four Android tablets (the main tablet is used to draw the robot's letter and to
acquire the children's demonstrations; more tablets have been used in some
studies, either to let the child input words to be written, or for the
experimenter to qualitatively annotate the interaction in a synchronized
fashion), and a central laptop running the machine learning algorithms, the
robot's handwriting gesture generation (based on the NaoQI inverse kinematics
library) and the high-level control of the activity (relying on {\sc
pyRobots}~\cite{lemaignan2015pyrobots} and a custom finite state machine).

@@

The system has been deployed and tested in several situations: in three
different schools (more than 70 children aged 5 to 8; relatively short duration
interactions), in an occupational therapy clinic (8 children; each
interacting several hours with the system), and during two case studies that
each lasted several weeks. Table~\ref{studies} gives an overview of these
studies.

\begin{table}[ht!]
\centering
\caption{\small Field studies conducted within the project}
\label{studies}
\footnotesize
\begin{tabular}{@{}lp{4cm}p{2.2cm}p{2cm}l@{}}
\toprule
{\bf Study}        & {\bf Type}                               & {\bf Avg. duration}    & {\bf \# Children}                     & {\bf Ages} \\ \midrule
{\it School A (1)} & Unstructured group interaction at school & 16 min/group           & 4 $\times$ 8 children                 & 6-7        \\
{\it School B}     & Individual/Pair interaction at school    & 11 min/group           & 7 (individual) + 7 $\times$ 2 (pairs) & 7-8        \\
{\it School C}     & Pair interaction at school               & 26 min/group           & 7 $\times$ 2                          & 5-6        \\
{\it School A (2)} & Individual interaction at school         & 20 min                 & 6                                     & 5-6        \\ \midrule
{\it Clinic}       & Individual interaction at clinic         & 3 sessions $\times$ 1h & 8                                     & 6-8        \\ \midrule 
{\it Vincent}      & Case-study                               & 4 weeks $\times$ 1.5h  & 1                                     & 6          \\
{\it Thomas}       & Case-study                               & 4 weeks $\times$ 1h    & 1                                     & 5          \\ \bottomrule
\end{tabular}
\end{table}

We report hereafter the main design choices and results for each of these
studies and experiments. The interested reader can find supplementary details
in~\cite{jacq2016building, hood2015when}.

@@

\begin{figure}
    \centering
    \includegraphics[height=3.9cm]{schools}
    \includegraphics[height=3.9cm]{schools2}
    \caption{\small The first field studies were focused on technical validation, with
    more than 70 pupils interacting with the robot over short periods (between
    10 and 25 minutes), either alone or in small groups.}
    \label{fig:schools}
\end{figure}

@@

\paragraph{System Validation}

@@

Children's recognition that the robot is writing by itself is critical for our
approach to be effective. When asked, no child indicated that they did not
believe that the robot was writing by itself: that it was simply miming the 
actions in the air as the tablet displayed the letters. There were, at times, questions
about the robot's writing without a pen at the beginning of the interaction, but when
advised that the robot ``tells the tablet what it wants to write,'' this was
accepted by the children.  In addition, teachers interviewed for their feedback on
the system advised that children are asked to draw letters in the air in a
similar manner as part of their handwriting education. The behavior is hence
not unfamiliar to children.

@@

The literature suggests that 20 handwriting practice sessions is found to be the
minimum to demonstrate effective results in handwriting
remediation~\cite{Hoy2011}. This highlights the necessity to sustain a
child-robot engagement over the long-term if we want to achieve measurable
learning gains.

Factually, the children engaged into the teaching activity: in the
\textit{School B} study, for instance, they demonstrated an average of 10.9
demonstration letters (SD = 4.4) for an average session duration of 11 minutes.
In 9 out of the 14 sessions (64\%), the robot received demonstration letters
even \emph{after} reaching the final stage of the interaction, suggesting an
intrinsic motivation to further engage in the interaction.

@@

We also conducted a quantitative assessment of the engagement levels of the
children. Table~\ref{tab:results-with-me-ness} reports the levels of
\emph{with-me-ness} of the children during the second study at school A.
\emph{With-me-ness} is a quantifiable precursor of engagement: it measures the
percentage of time spent by the child focusing on the task at hand. This metric
was first devised in the context of computer-supported
learning~\cite{sharma2014me}, and we have previously studied its applicability
to human-robot interaction and formalized the exact methodology
in~\cite{lemaignan2016realtime}. We compute it by first estimating the focus of
attention of the child over the course of the interaction (using real-time 6D
head pose estimation); then, by matching this measured focus of attention
with the \emph{expected} attentional targets. For instance, when the child is supposed
to write a demonstration, we expect him to look at the tablet if he is actually
engaged into the task; when the robot is telling a story,
we expect the engaged child to look at the robot, etc. The resulting values
(Table~\ref{tab:results-with-me-ness}) represent therefore the percentage of
interaction where each of the children were actually focusing on the task\footnote{Note that
with-me-ness is a metric that allows real-time computation by the robot
itself: while we did not yet make use of it, it does in principle allow the
robot to detect on-line possible disengagements, and eventually address them
by adapting its physical behavior, switching to different activities, etc.}.

\begin{table}[h!]
    \centering
    \caption{\textbf{Levels of with-me-ness}. Percentage of interaction time
        during which the child was effectively focusing his/her attention on the
        task. The six children are those from the second study at School A.
        Interaction duration: M = 19.6 min, SD = 1.58. Results
        taken from~\cite{lemaignan2016realtime}.}

    \begin{tabular}{p{1cm}cccccccc}
        \toprule
        \bf Child & 1 & 2 & 3 & 4 & 5 & 6 & {\bf M} & {\it SD} \\
        \midrule
        $\mathcal{W}$ & 79.4\% & 81.6\%  & 90.5\% & 87.9\% & 90.7\% & 80.9\% & {\bf 85.2\%} & {\it 5.1} \\ 
        \bottomrule
    \end{tabular}
    \label{tab:results-with-me-ness}
\end{table}

@@

The other three experiments (the group study at the clinic and the two
case-studies) provide further qualitative evidence of engagement over longer
interaction periods. In particular, as reported hereafter, the two 4-week long
case studies that have been conducted so far show that our system can sustain
children's engagement over durations (5 hours) that are closer to what is
expected to have an impact in a real therapeutic context.

@@

\begin{inparaenum}[\itshape 1\upshape)]
This experiment was conducted to study \item how easily children with actual
deficits take on the role of a teacher, and \item if they adopt this role.
Measuring actual handwriting improvements was not a primary goal of this
experiment.
\end{inparaenum}

The experiment took place at an occupational therapy clinic in
Normandy, France. Eight children participated, selected by the occupational
therapist based on their age and type of deficit (all related to handwriting).
Valérie (female, 7 years old), Antoine (male, 6.5) and Johan (male, 7) are under
the direct care of the occupational therapist. Émilien (male, 8) and Mathieu
(male, 7) are repeating their school year because of writing difficulties. Marie
(female, 6) and Adèle (female, 8) are both ranked at the bottom of their
respective classes in writing activities. Nicolas (male, 7) is under the care of
a neurologist, and has been diagnosed with specific language impairment. Given
their age and school year, all of these children would be expected to know how to
correctly shape cursive letters.

Over a period of two weeks, each child attended three times a one hour long
session (except for Marie and Adèle who only attended one session). The
experimenter's role was limited to the explanation of the task and the basic
tablet usage. For this experiment, the children were provided with two tablets:
one to choose the words (or letters) to teach to the robot, the other one to
write, as in the validation studies. We also provided paper-based templates of
the cursive letters, were the children to ask for them.

We only provided the children with minimal explanations on the task (they would
have to help the robot to improve its writing style), so as to assess if and how
children would naturally take on the role of the teacher. We additionally assessed
how seriously they engage into helping the robot through two additional buttons
on the tablet: a green ``thumbs up'' and a red ``thumbs down''. The children
were told to freely use them to evaluate the robot's improvements (``thumb up''
to give positive feedback, ``thumb down'' to convey negative feedback). Our
assessment builds on the hypothesis that the more the child provides feedback to
the robot, the more they assume the role of a teacher. Then, by correlating the
feedback with the actual performance of the robot, we can measure to what extent
the children are adopting their teacher role: if their feedback does correlate
with the actual performance of the robot, the child has likely successfully
adopted the teacher role.

@@

\begin{table}
    \centering
    \begin{tabular}{ccccl}
        \toprule
        \bf Child      & \bf \# Demos/hour & \bf \# Positive & \bf \# Negative & $r$ \\ \midrule
        \emph{Valérie} & 42           & 24              & 6               & 0.25 \small\tt **  \\ 
        \emph{Émilien} & 74           & 20              & 9               & 0.06 \small\it ns  \\
        \emph{Mathieu} & 43           & 10              & 3               & 0.23 \small\tt **  \\
        \emph{Nicolas} & 38           & 16              & 4               & 0.31 \small\tt *** \\
        \emph{Johan}   & 32           & 10              & 5               & 0.10 \small\it ns  \\
        \emph{Antoine} & 27           & 10              & 3               & 0.20 \small\tt *   \\
        \emph{Adèle}   & 35           & 4               & 2               & 0.28 \small\tt *   \\
        \emph{Marie}   & 40           & 5               & 1               & -0.02 \small\it ns \\ \bottomrule
    \end{tabular}
    \caption{\small Feedback from the children to the robot. \emph{\#Demos}
        denotes the average number of demonstrations per hour provided by the children;
        \emph{\#Positive} and \emph{\#Negative} the total number of positive \resp
        negative feedbacks they provided. $r$ is the correlation coefficient
        between the feedback provided by the children and the performance of the
        robot. Results taken from~\cite{jacq2016building}.}

    \label{table:scores}
\end{table}

All children maintained their engagement during the entirety of the sessions. They
provided on average 42 demonstrations per session. All children used the
feedback buttons (in total, 99 ``thumbs up'' and 33 ``thumbs down'' were recorded,
see Table~\ref{table:scores}). This indicates that they are all able to engage
into playing the role of a teacher.

To study the correlation between the children's feedback and the actual
improvements of the robot, we estimate the robot's progress as the difference
between an initial score (euclidian distance between the shape drawn by
the robot at its first attempt and the shape of the reference letter) and the
robot's score at the current round of demonstrations. We then correlate this
progress to the positive or negative feedback provided by the children (details
of the method are presented in~\cite{jacq2016building}).

We find that the feedback of five out of the eight children does significantly
correlate with the actual performance of the robot (Table~\ref{table:scores}).
This indicates that these children are effectively taking on the teacher's
role and are seriously providing feedback to the robot. The observation of
the three remaining children reveal a variety of behaviors (for instance, one
was actually rating how ``nice'' the robot is, and another one had rather
``artistic'' writing style preferences that were independent of the actual
legibility of the robot), but none of these children adopted a playful-only
behavior toward the robot.

To summarize, this experiment shows that children with actual handwriting
impairments do accept well the robot, can commit themselves into long
interactions, and do adopt the role of the teacher (as evidenced by both the
number of writing demonstrations they provided and their self-inclination to
give feedback to the robot).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\linewidth]{diego}
    \caption{\small Vincent correcting {\sc nao}'s attempt by rewriting the
        whole word. Empty boxes are drawn on the screen to serve as a template for the child
        and to make letter segmentation more robust.}
    \label{fig:diego}
\end{figure}

@@

Despite his attention deficit, Thomas was able to remain engaged in the activity
during more than forty minutes in each session (a long time for a five year
old). In total, 55 allographs out of 82 demonstrated by the child were
acceptable considering our threshold (with a progressive improvement from 13 out
of 28 in the first session up to 26 out of 29 in the last session).

@@

In total, more than 80 children have interacted with the system, for a total
duration of more than 38 hours, in multiple experimental configurations
(individual interaction \vs pairs \vs groups; at school, in the lab, at a
occupational therapy clinic; short interactions \vs long, repeated interactions; 5 to 8 year old
children). We summarize hereafter the main findings from these experiments and
discuss some of the critical points of our approach.

>>>>

\documentclass[twocolumn]{article}
%\documentclass{article}

@@

\subsection*{Learning by Teaching Handwriting}

@@

Robots have been used as teachers or social partners to promote children's
learning in a range of contexts, most commonly related to language
skills~\cite{han2010robot}, and less often to physical skills (such as
calligraphy~\cite{Matsui2013}). Looking at the converse (humans \emph{teaching}
robots), Werfel notes in~\cite{Werfel2014} that most of the work focuses on the
robot's benefits (in terms of language~\cite{Saunders2010} or
physical~\cite{Mulling2013} skills, for example) rather than the learning
experienced by the human tutor themselves.  Our work concentrates on this latter
aspect: by demonstrating handwriting to a robot, we aim at improving the
\emph{child's} performance.

@@

\subsection*{Agency and Commitment}

@@

For these two reasons, our approach is to be distinguished from previous works
in educational robotics. Most of these do not consider the agency induced by the
robot beyond its motivational aspect (playing with an interactive, partially
autonomous device induces excitement -- at
least, on the short-term).  In our case, the
role of agency is stronger: it induces meta-cognition (``I am interacting with
an agent, so I need to reflect on how to best teach him'') which is beneficial
for the learning process; it also induces a protégé effect (``I want my
robot-agent to succeed!'') which supports the commitment of the child into the
interaction, also for longer periods of time.

\vspace{2em}
These initial considerations can be turned into a set of research questions: how
to develop a fully autonomous robotic system that produces believable
handwriting, foster interaction and adequately learn from children's demonstration?
Can such a system be accepted by children and practitioners alike, and
eventually be integrated in existing educative and therapeutic practises?
Can we build and evidence interactions that sustains children's engagement over extended
periods of time? And finally, is this novel approach effective? Do we evidence
actual handwriting improvements by children facing difficulties?

We attempt to provide initial insights on these questions in the remaining
sections of the article. We first provide an overview of the robotic system and
the interaction it induces in Section~\ref{impl}. We present there the
machine-learning techniques that allow the robot to learn from the children, as
well as the actual implementation of the system on a {\sc nao} robot.

@@

Figure~\ref{experimental_setup} illustrates our general experimental setup: a
face-to-face child-robot interaction with an autonomous Aldebran's {\sc nao}
robot. A tactile tablet is used for both the robot and the
child to write. During a typical round, the child requests the robot to write
something (a single letter, a number or a full word), and pushes the tablet
towards the robot. The robot writes on the tablet by gesturing the writing in
the air,  the letters being actually drawn by the tablet application itself. The
child then pulls back the tablet, corrects the robot's attempt by writing
herself on top of or next to the robot's writing (see Figure~\ref{fig:diego}),
and sends her demonstration to the robot by pressing a small button on
the tablet. The robot learns from this demonstration and tries again.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.9\columnwidth]{experimental_setup}
    \caption{\small Our experimental setup: the robot writes on a tactile tablet, the child then
        corrects the robot by overwriting its letters with a stylus. An adult
        (either a therapist or an experimenter) guides the work
        (prompting, turn taking, etc.). For some studies, a second tablet and an
        additional camera (dashed) are employed.}

@@

@@

The main insight for both the generation and the learning of letters is to
reason about the shape of letters in their eigenspace, instead of the natural
cartesian space. The eigenspace of each letter is spanned by the first $n$
eigenvectors (in our experiments, $3 < n < 6$) of the covariance matrix
generated from a standard dataset of adult letters (the UJI Pen Characters 2
dataset~\cite{Llorens2008}).  Our method, based on a Principle Component
Analysis (PCA), is presented in~\cite{hood2015when}.

New letters are generated by varying the weightings in linear combinations of
these eigenvectors, with distortions that are actually plausible handwriting errors: 
they are exaggerations of variations of writing styles that naturally occur amongst
adult writers.  Figure~\ref{fig:sampleLetters} shows examples of deformed ``g''
generated with such a technique.

\begin{figure}
    \centering
    \includegraphics[width=0.9\linewidth]{cowriter-g}
    \caption{\small \label{fig:sampleLetters} \textbf{Generating bad letters}:
        effect of independently varying the weightings (columns) of the first five eigenvectors 
        (rows) of the shape model of ``g''. Examples A, B, C, D illustrate how the
        PCA-based approach allows for the \emph{automatic} generation of letters whose
        errors are typical of children handwriting: A has too large a bottom
        loop, B has a wide top loop, the bottom loop of C is not correctly
        closed, the top loop of D is not closed, etc.}

@@

The same shape model can also be used to \emph{classify} demonstrations,
\emph{assess} their quality and \emph{learn} from them. Figure~\ref{fig:h} shows
for example nine allographs of ``h'' written by a 6 years old child, along with
the reference shape. By projecting each of the
demonstrations onto the eigenspace of ``h'' (Figure~\ref{fig:h}b), we observe
that:

@@

\begin{figure*}[ht!]
    \centering
    \subfigure[Nine allographs of the cursive ``h'', next to the reference]{
        \includegraphics[width=0.45\linewidth]{h}
    }
    \subfigure[The same samples, normalized and projected in the eigenspace spanned by the
        first 3 eigenvectors: clusters arise, that actually match writing styles.]{
        \includegraphics[width=0.45\linewidth]{eigenspace-uniformization}
    }

    \caption{\small Projecting demonstrated letters onto the eigenspace
    generated from the reference dataset effectively clusters the samples
    according to their topological similarity. Allographs that are similar to
    the reference are close to it in the eigenspace.}
    \label{fig:h}
\end{figure*}

@@

The overall architecture of the system is therefore
spread over several devices: the {\sc nao} robot itself, that we address via
both a ROS API and the Aldebaran-provided NaoQI API, one to
four Android tablets (the main tablet is used to draw the robot's letter and to
acquire the children's demonstrations; more tablets have been used in some
studies, either to let the child input words to be written, or for the
experimenter to qualitatively annotate the interaction in a synchronized
fashion), and a central laptop running the machine learning algorithms, the
robot's handwriting gesture generation (based on the NaoQI inverse kinematics
library) and the high-level control of the activity (relying on {\sc
pyRobots}~\cite{lemaignan2015pyrobots} and a custom finite state machine).

@@

The system has been deployed and tested in several situations: in three
different schools (more than 70 children aged 5 to 8; relatively short duration
interactions), in an occupational therapy clinic (8 children, each
interacting several hours with the system), and during two case studies that
each lasted several weeks.
We report hereafter the main design choices and results for each of these
studies and experiments. The interested reader can find supplementary details
in~\cite{jacq2016building, hood2015when}.

@@

\begin{figure}
    \centering
    \includegraphics[height=5cm]{schools2}
    \caption{\small The first field studies were focused on technical validation, with
    more than 70 pupils interacting with the robot over short periods (between
    10 and 25 minutes), either alone or in small groups.}
    \label{fig:schools}
\end{figure}

@@

\paragraph{System validation}

@@

Children's recognition that the robot is writing by itself is critical for our
approach to be effective. When asked, no child indicated that they did not
believe that the robot was writing by itself. There were, at times, questions
about the robot's writing without a pen at the beginning of the interaction, but when
advised that the robot ``tells the tablet what it wants to write,'' this was
accepted by the children.

@@

Factually, the children engaged into the teaching activity: in the
study at school B, for instance, they demonstrated an average of 10.9
demonstration letters (SD = 4.4) for an average session duration of 11 minutes.
In 9 out of the 14 sessions (64\%), the robot received demonstration letters
even \emph{after} reaching the final stage of the interaction, suggesting an
intrinsic motivation to further engage in the interaction.

@@

We also conducted a quantitative assessment of the engagement levels of the
children. Table~\ref{tab:results-with-me-ness} reports the levels of
\emph{with-me-ness} of the children during the second study at school A.
\emph{With-me-ness} is a quantifiable precursor of engagement: it measures the
percentage of time spent by the child focusing on the task at hand~\cite{lemaignan2016realtime}. We compute it by first estimating the focus of
attention of the child over the course of the interaction (using real-time 6D
head pose estimation); then, by matching this measured focus of attention
with the \emph{expected} attentional targets. For instance, when the child is supposed
to write a demonstration, we expect her to look at the tablet if she is actually
engaged into the task; when the robot is telling a story,
we expect the engaged child to look at the robot, etc.

\begin{table*}[h!]
    \centering
    \caption{\textbf{Levels of with-me-ness}. Percentage of interaction time
        during which the child was effectively focusing her attention on the
        task. The six children are those from the second study at School A.
        Interaction duration: M = 19.6 min, SD = 1.58. Results
        taken from~\cite{lemaignan2016realtime}.}

    \begin{tabular}{p{1cm}cccccccc}
        \toprule
        \bf Child & 1 & 2 & 3 & 4 & 5 & 6 & {\bf M} & {\it SD} \\
        \midrule
        $\mathcal{W}$ & 79.4\% & 81.6\%  & 90.5\% & 87.9\% & 90.7\% & 80.9\% & {\bf 85.2\%} & {\it 5.1} \\ 
        \bottomrule
    \end{tabular}
    \label{tab:results-with-me-ness}
\end{table*}

@@

@@

This experiment was conducted to study how easily children with actual
deficits take on the role of a teacher, and to what extent they adopt this role
(measuring actual handwriting improvements was not a primary goal of this
experiment).

The experiment took place at an occupational therapy clinic in Normandy, France.
Eight children participated, selected by the occupational therapist based on
their age and type of deficit (all related to handwriting).  Given their age (6
to 8 years old) and school year, all of these children would be expected to know
how to correctly shape cursive letters. Each child attended three times a one
hour long session spread over two weeks (except for two of them who only
attended one session). The experimenter's role was limited to the explanation of
the task and the basic tablet usage. The children were provided with two
tablets: one to choose the words (or letters) to teach to the robot, one to
write.

We only provided the children with minimal explanations on the task (they would
have to help the robot to improve its writing style), so as to assess if and how
children would naturally take on the role of the teacher. We additionally
assessed how seriously they engage into helping the robot through two additional
buttons on the tablet: a green ``thumbs up'' and a red ``thumbs down''. The
children were told to freely use them to evaluate the robot's improvements
(``thumb up'' to give positive feedback, ``thumb down'' to give negative
feedback). Our assessment builds on the hypothesis that the more the child
provides feedback to the robot, the more they assume the role of a teacher.
Then, by correlating the feedback with the actual performance of the robot, we
can measure to what extent the children are adopting their teacher role: if
their feedback does correlate with the actual performance of the robot, the
child has likely successfully adopted the teacher role.

@@

All children maintained their engagement during the whole sessions. They
provided on average 42 demonstrations per session. All children used the
feedback buttons (in total, 99 ``thumbs up'' and 33 ``thumbs down'' were
recorded). This indicates that they are all able to engage into playing the role
of a teacher.

To study the correlation between the children's feedback and the actual
improvements of the robot, we estimate the robot's progress as the difference
between an initial score (euclidian distance between the shape drawn by the
robot at its first attempt and the shape of the reference letter) and the
robot's score at the current round of demonstrations. We then correlate this
progress to the positive or negative feedback provided by the children (details
of the method, as well as the complete results, are presented
in~\cite{jacq2016building}).

We find that the feedback of five out of the eight children does significantly
correlate with the actual performance of the robot.  This indicates that these
children are effectively taking on the teacher's role and are seriously
providing feedback to the robot. The observation of the three remaining children
reveal a variety of behaviors (for instance, one was actually rating how
``nice'' the robot is, and another one had rather ``artistic'' writing style
preferences that were independent of the actual legibility of the robot), but
none of these children adopted a playful-only behavior toward the robot.

To summarize, this experiment shows that children with actual handwriting
impairments do accept well the robot, can commit themselves into long
interactions, and do adopt the role of the teacher (as evidenced by both the
number of writing demonstrations they provided and their self-inclination to
give adequate feedback to the robot).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.8\linewidth]{diego}
    \caption{\small Vincent correcting {\sc nao}'s attempt by rewriting the
        whole word. Empty boxes are drawn on the screen to serve as a template for the child
        and to make letter segmentation more robust.}
    \label{fig:diego}
\end{figure}

@@

Despite his attention deficit, Thomas was able to remain engaged in the activity
during more than forty minutes in each session (a long time for a five years
old). In total, 55 allographs out of 82 demonstrated by the child were
acceptable considering our threshold (with a progressive improvement from 13 out
of 28 in the first session up to 26 out of 29 in the last session).

@@

In total, more than 80 children have interacted with the system, for a total
duration of more than 38 hours, in multiple experimental configurations
(individual interaction \vs pairs \vs groups; at school, in the lab, at an
occupational therapy clinic; short interactions \vs long, repeated interactions;
5 to 8 years old
children). We summarize hereafter the main findings from these experiments and
discuss some of the critical points of our approach.

\paragraph{Technical Assessment} We set ourselves the challenge of developing an
autonomous robotic system able to perform handwriting tasks with children in
real world environments. The system performed generally well: the children had no
issues interacting with the robot, the robot experienced relatively few crashes,
and those crashes did not significantly impair the interactions. The system
has furthermore proved robust enough to conduct several hour-long experiments.
However, we also found that the level of expertise required to deploy and
operate our system still makes the presence of an experimenter mandatory at all
time. Non-trivial technical developments may be required to reach a level of
usability suitable for a broader, non-expert, audience.
